### 2020-06-23 chapters/sota.tex
SOTA improvements

A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}. In this Chapter, we discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems. We present a wide analysis of several categories of algorithms that seem viable to build the desired system.

@@

\section{Outlier detection in time-series data} \label{sec:outliers}
Blázquez-García et. al review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented in \textit{"A review on outlier/anomaly detection in time-series data"} by Blázquez-García et. al due to its focus on time-series data.

A time-series is a collection of data points sorted by time. Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. In summary, outlier detection refers to the process of finding anomalies in data. An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

The authors evaluate an outlier detection method on three different axes: the input data type, the outlier type to detect and the uni or multivariate nature of the method. Figure \ref{fig:outlier-taxonomy} presented by the authors summarizes these categories.

@@

Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy}).

@@

The univariate point outlier detection methods based on models are the most common. For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value. If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier. Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if
\begin{equation*}
    |x\textsubscript{t}-e\textsubscript{t}|>\alpha
\end{equation*}
Model-based point outlier detection methods can be one of two types: estimation or prediction models. The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values. Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can not.

@@

Yang et. al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios. The authors propose Extra-N, an algorithm that detects point outliers based on the number of neighbors it has. The authors make use of the \textit{“predictability"} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.

The authors want to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has, rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors. The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers. The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors. The authors introduce the concept of \textit{"lifetime neighbor counts"} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where he exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window. The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Applying the \textit{predictability} property makes it possible to compute the lifespan of all points. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, 
\[ \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\} \]
The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.

\subsubsection*{Method classification}
Extra-N can take both uni or multivariate time-series as input data. However, it does not relate multiple time-series and does not take into account possible dependencies between different time-dependent variables. In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method accepts uni or multivariate time-series as input but is of univariate nature. Last but not least, Extra-N aims to detect univariate point outliers. These point outliers are identified based on the density of the region or neighborhood they belong too. Hence, we say Extra-N is a density-based univariate point outlier detection method (Figure \ref{fig:point-outlier-taxonomy}).

@@

\subsubsection*{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

\subsubsection{DeepAnT}
Munir et. al present DeepAnT, a deep learning-based approach for the detection of anomalies in time series data \cite{Munir-DeepAnT}, that uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series. It is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known. It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.

@@

\subsection{Univariate subsequence outlier detection}
Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}). Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.

@@

Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series to find out the most unusual ones. Each subsequence is compared to the others using a distance metric, for instance, the Euclidean distance. The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in quadratic time complexity. These methods find the most unusual subsequence in a time-series. However, as the authors put it, these types of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.

@@

Prediction model-based univariate subsequence outlier detection methods assume that the reference or normality period is composed of past events. Similarly to univariate point outlier detection methods based on prediction models, these models train on past data and make predictions of the future. Subsequences that do not resemble the prediction made by the model are classified as outliers.

@@

Lastly, we have information theory based methods for univariate subsequence outlier detection, which according to the authors \textit{"are closely related to the frequency-based methods"}. This type of method not only measures the frequency of subsequences but relates it with the information carried by the subsequence. Mathematically, a subsequence \textit{S} will be an outlier if

@@

with \textit{F(S)} as the frequency of subsequence S, \textit{I(S)} as the information carried by S and $\alpha$ as a given threshold. According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}. The authors explain that if \textit{F(S)} is large --- \textit{i.e.} S occurs frequently --- then \textit{I(S)} will be closer to 0.

\subsubsection{A two window paradigm algorithm for change detection}
Kifer et al. propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions \cite{Kifer-Detecting-Change}. The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}. Both windows are tuple-based, have size \textit{k} and the sliding window is a true sliding one (recall Section \ref{sec:windows} and the expression \ref{math:windows}). The reference window \textit{W\textsubscript{2}} works as a \textit{reference of normality} and contains the first \textit{k} points of the stream that occurred after the last detected change.

@@

Note that after a change is detected, the reference window is updated with the next \textit{k} tuples from the stream. As we have stated, we want a method that does not adapt to the data stream changes and instead keeps raising alerts until the sliding window data distribution returns to the one seen in the reference period. It is trivial to adapt this method to always use the same reference period. To do so, we simply keep the original reference window and don't update it after each alert.

@@

\subsubsection*{Method classification}
The method devised by Kifer et. al can take uni or multivariate input (if we use one pair of reference/target windows for each time-series) but performs a univariate analysis as it does not find correlations between time-dependent variables. This two-windowed method measures the difference between both windows, the reference and the sliding one. Furthermore, instead of detecting point outliers, the alerts produced indicate the entire window (or subsequence) shows anomalous behavior. Hence, according to the taxonomy presented in Figure \ref{fig:subsequence-outlier-taxonomy}, this method is a dissimilarity-based univariate subsequence outlier detection method.

@@

However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one. Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we may use a similar two-windowed model if we find lightweight and incrementally maintained aggregations over both windows.

\subsubsection{Automatic Model Monitoring for Data Streams}

ALSO a 2 WINDOW MODEL
\textcolor{red}{TODO: write this}

@@

The De-Amortized Banker’s Aggregator (DABA) \cite{Tangwongsan-DABA} is a SWAG algorithm that allows the computation of \textit{associative} aggregations. Quoting the authors: \textit{"Deamortization is a method that turns the average-case behavior into the worst-case behavior, usually by carefully spreading out expensive operations"}. Hence the name \textit{"De-Amortized"}, as the main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout run time.

@@

DABA uses two queues, \textit{vals} and \textit{aggs}, both of size \textit{n}, seen in Figure \ref{fig:daba-ds}, which was directly taken from \textit{"Low-Latency Sliding-Window Aggregation in Worst-Case Constant Time"} by Tangwongsan et. al. Both queues share six pointers --- \textit{F, L, R, A, B} and \textit{E} --- that point to the same position for each queue. These pointers are always ordered as follows: $\textit{F} \leq \textit{L} \leq \textit{R} \leq \textit{A} \leq \textit{B} \leq \textit{E}.$

@@

\textit{l\textsubscript{F}} is aggregated to the left for easier eviction. After an element is evicted, the new head of the queue \textit{aggs} --- pointed by \textit{F} --- already contains the aggregation value of all elements from \textit{vals} to its right side. Hence, evicting an element amounts to evicting from both \textit{vals} and \textit{aggs} queue. Similarly, \textit{l\textsubscript{B}} is aggregated to the right to facilitate insertion. Sub-lists \textit{l\textsubscript{L}}, \textit{l\textsubscript{R}} and \textit{l\textsubscript{A}} allow for incremental reversal of the right-aggregated \textit{l\textsubscript{B}} to the left-aggregated \textit{l\textsubscript{F}}. This incremental reversal is tantamount to adjusting pointers that causes shifts in sub-list boundaries. Thus, elements of \textit{aggs} may change from one sub-list to another and need to have its aggregation value recomputed.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/query-bloom.png}
      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

@@

\subsection*{False Positive rate}
% error bound related with parameters
The false-positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}. It is assumed that the hash functions produce an independent and random index value for each element. Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula.

The probability of one arbitrary bit being set to 1 is
\begin{equation}
    \frac{1}{m}
\end{equation}
because there are \textit{m} bits. Likewise, the probability of one bit not being set to 1 is
\begin{equation}
    1-\frac{1}{m}
\end{equation}

Mullin states the probability that an arbitrary bit is not set to 1 after one element insertion and corresponding \textit{k} bit updates is
\begin{equation}
    (1-\frac{1}{m})^\textit{k}
\end{equation}

For \textit{n} insertions, the probability a bit is not set to 1 is
\begin{equation}
    (1-\frac{1}{m})^\textit{kn}
\end{equation}

Therefore, the probability that an arbitrary bit is set to 1 after \textit{n} insertions is 
\begin{equation}
    1-(1-\frac{1}{m})^\textit{kn}
\end{equation}

@@

This false positive rate can be approximated by 
\begin{equation}
    (1-e^\textit{-kn/m})^\textit{k}
\end{equation}

\subsection*{Optimal number of hash functions}
We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate. In order to do so, we must first allocate \textit{m} bits for the Bloom filter, choosing \textit{m} based on available memory. The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by
\begin{equation}
    \frac{m}{n}ln(2)
\end{equation}

@@

Note the time complexity does not at all depend on the number of elements in it and that \textit{k} will be a rather small constant. Hence, the Bloom filter is said to have constant time complexity.

@@

% data structures
\subsection*{Data structure}
A CMS is represented by a two-dimensional array (or matrix) with width \textit{w} and depth \textit{d}. A CMS will make use of \textit{d} hash functions --- \textit{i.e.} each associated with one row of the matrix. Each hash function will map elements to a position between 1 and \textit{w} --- \textit{i.e.} the column of the two-dimensional array. Initially, all of the matrix positions are 0. Figure \ref{fig:initial-cms} represents the two-dimensional array used by the CMS and corresponding dimensions in its initial state.

@@

% algorithm for insertion
\subsection*{Element Insertion}
When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, hash the element using that row's hash function to obtain \textit{j}. Lastly, for all obtained pairs of \textit{i} and \textit{j}, increment matrix cells \textit{(i, j)} value by 1, as seen in Figure \ref{fig:cms}. If elements have different weights these can be added instead of incrementing one unit.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/insertion-cms.png}
      \caption[Count-Min Sketch insertion]{CMS mapping element \textit{el} using \textit{d}=4 hash functions to determine the column position for each row and increment it}
      \label{fig:cms}
    \end{center}
\end{figure}

% algorithm for query
\subsection*{Frequency Query}
Retrieving the minimum frequency of an element \textit{el} can be done by taking the minimum value of all row counts for \textit{el}. Mathematically, the frequency estimate is the minimum value of the set of all row values, for each row \textit{i} of the total \textit{d} rows, given by
\[ min \{matrix[i][\textit{h\textsubscript{i}}(el)]\} \]

@@

\subsection*{Element Insertion}
In HyperLogLog, new elements are hashed into a binary string. The first \textit{n} bits of the hashed representation will map to one of the \textit{m}=$2^n$ buckets. Effectively speaking, the first \textit{n} bits work as a \textit{bucket ID}. The bit-pattern observable \textit{P} --- the position of the leftmost 1-bit --- is counted from the binary string discarding the first \textit{n} bits. The observable \textit{P} is stored in the pre-determined bucket if it is greater than the value the bucket contains.

Figure \ref{fig:hll-insert} demonstrates such procedure for a new element \textit{el}. The hash representation of \textit{el}, using \textit{n}=2 bits, maps to bucket \textit{10}. Discarding the first \textit{n}=2 bits and counting from 1, the position of the leftmost 1-bit is \textit{P}=4. The previous value of bucket \textit{10} was 0 and because \textit{P}=4 is greater than 0 the bucket value is updated.

>>>>

A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}. In this Chapter, we discuss the most relevant work for the problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems, presented in Chapter \ref{chap:intro}. We present a wide analysis of several categories of algorithms that seem viable to build the desired system.

@@

\section{Outlier detection in time-series data} \label{sec:outliers}
Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.

In this work, the authors define a time-series as a collection of data points sorted by time; and outliers as observations that do not follow the expected behavior. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is “an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism" \cite{Hawkins-Outliers}. Furthermore, Blázquez-García \emph{et al.} allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. In summary, outlier detection refers to the process of finding anomalies in data. An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

The authors evaluate an outlier detection method on three different axes: the input data type, the outlier type to detect and the uni or multivariate nature of the method. Figure \ref{fig:outlier-taxonomy} presented by the authors in \cite{Blazquez-Garcia-Review-Anomaly-Detection} summarizes these categories.

@@

Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy} \cite{Blazquez-Garcia-Review-Anomaly-Detection}) .

@@

The univariate point outlier detection methods based on models are the most common. For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value. If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier. Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if:
\begin{equation*}
    |x\textsubscript{t}-e\textsubscript{t}|>\alpha
\end{equation*}
Model-based point outlier detection methods can be one of two types: estimation or prediction models. The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values. Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can only be used in batch.

@@

Yang et. al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios. The authors propose \emph{Extra-N}, an algorithm that detects point outliers based on the number of neighbors it has. The authors make use of the \emph{predictability} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.

The authors of Extra-N want to maintain for each data point $p\textsubscript{i}$ the number of neighbors it has, rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors. The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers. The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors. The authors introduce the concept of \textit{lifetime neighbor counts} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where it exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window. Consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors: \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, 
\[ p\textsubscript{i}.lt\_cnt = \{W\textsubscript{i}: 3, W\textsubscript{i+1}: 2, W\textsubscript{i+2}: 0\} \]
The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.

\subsubsection*{Method classification}
Extra-N can take both uni or multivariate time-series as input data. However, it does not relate multiple time-series and it does not take into account possible dependencies between different time-dependent variables. In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method accepts uni or multivariate time-series as input but is of univariate nature. Last but not least, Extra-N aims to detect univariate point outliers. These point outliers are identified based on the density of the region or neighborhood they belong too. Hence, we say Extra-N is a density-based univariate point outlier detection method (Figure \ref{fig:point-outlier-taxonomy}).

@@

\subsubsection*{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable to our goals. In fact, we need a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, leaving the user confused on why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

\subsubsection{DeepAnT}
Munir et. al present DeepAnT, a deep learning-based approach for the detection of anomalies in time-series data \cite{Munir-DeepAnT}. DeepAnT uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series. This method is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known. It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.

@@

\subsection{Univariate subsequence outlier detection} \label{sec:uni-sub-out}
Similarly to point outlier detection methods, Blázquez-García \emph{et al.} \cite{Blazquez-Garcia-Review-Anomaly-Detection} split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}). Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.

@@

Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series to find out the most unusual ones. Each subsequence is compared to the others using a distance metric, for instance, the Euclidean distance. The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in quadratic time complexity. These methods find the most unusual subsequence in a time-series. However, as the authors put it, this type of methods lack a reference or normality state and hence do not accurately report outliers.

@@

Prediction model-based univariate subsequence outlier detection methods assume that the reference or normality period is composed of past events. Similarly to univariate point outlier detection methods based on prediction models, these models train on past data and make predictions of the future. Subsequences that deviate from the model's prediction are classified as outliers.

@@

Lastly, we have information theory based methods for univariate subsequence outlier detection, which according to the authors \textit{"are closely related to the frequency-based methods"}. This type of method not only measures the frequency of subsequences but relates it with the information carried by the subsequence. Mathematically, a subsequence $S$ will be an outlier if

@@

with $F(S)$ as the frequency of subsequence $S$, $I(S)$ as the information carried by $S$ and $\alpha$ as a given threshold. According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}. The authors explain that if \textit{F(S)} is large --- \textit{i.e.} S occurs frequently --- then \textit{I(S)} will be closer to 0.

\subsubsection{A two window paradigm algorithm for change detection}
Kifer \emph{et al.} propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions \cite{Kifer-Detecting-Change}. The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}. Both windows are tuple-based, have size \textit{k} and the sliding window is a true sliding one (recall Section \ref{sec:windows} and the expression \ref{math:windows}). The reference window \textit{W\textsubscript{2}} works as a \textit{reference of normality} and contains the first \textit{k} points of the stream that occurred after the last detected change.

@@

Note that after a change is detected, the reference window is updated with the next \textit{k} tuples from the stream. As we have stated, we want a method that does not adapt to the data stream changes and instead keeps raising alerts until the sliding window data distribution returns to the one seen in the reference period. It is trivial to adapt this method to always use the same reference period. To do so, we simply keep the original reference window and do not update it after each alert.

@@

\subsubsection*{Method classification}
The method devised by Kifer \emph{et al.} can take uni or multivariate input (if we use one pair of reference/target windows for each time-series) but performs a univariate analysis as it does not find correlations between time-dependent variables. This two-windowed method measures the difference between both windows, the reference and the sliding one. Furthermore, instead of detecting point outliers, the alerts produced indicate the entire window (or subsequence) shows anomalous behavior. Hence, according to the taxonomy presented in Figure \ref{fig:subsequence-outlier-taxonomy}, this method is a dissimilarity-based univariate subsequence outlier detection method.

@@

However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one. Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we use a similar two-windowed model, but use lightweight and incrementally maintainable aggregations over both windows.

\subsubsection{Automatic Model Monitoring for Data Streams}
ALSO a 2 WINDOW MODEL
\textcolor{red}{TODO: write this}

@@

The De-Amortized Banker’s Aggregator (DABA) \cite{Tangwongsan-DABA} is a SWAG algorithm that allows the computation of \textit{associative} aggregations. This method carefully handles expensive operations and spreads them along the execution. As a resut, the authors are able to turn the average-case behavior into the worst-case behavior. This is commonly known as \textit{de-amortization}, hence the name. The main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout run time.

@@

DABA uses two queues, \textit{vals} and \textit{aggs}, both of size \textit{n}, seen in Figure \ref{fig:daba-ds}, which was directly taken from \textit{"Low-Latency Sliding-Window Aggregation in Worst-Case Constant Time"} by Tangwongsan \emph{et al.} Both queues share six pointers --- \textit{F, L, R, A, B} and \textit{E} --- which point to the same position for each queue. These pointers are always ordered as follows: $F \leq L \leq R \leq A \leq B \leq E$.

@@

Figure \ref{fig:daba-ds} shows sub-lists \textit{l\textsubscript{F}}, \textit{l\textsubscript{B}}, \textit{l\textsubscript{L}}, \textit{l\textsubscript{R}} and \textit{l\textsubscript{A}}. Sub-list \textit{l\textsubscript{F}} is aggregated to the left for easier eviction. After an element is evicted, the new head of the queue \textit{aggs} --- pointed by \textit{F} --- already contains the aggregation value of all elements from \textit{vals} to its right side. Hence, evicting an element amounts to evicting from both \textit{vals} and \textit{aggs} queue. Similarly, \textit{l\textsubscript{B}} is aggregated to the right to facilitate insertion. Sub-lists \textit{l\textsubscript{L}}, \textit{l\textsubscript{R}} and \textit{l\textsubscript{A}} allow for incremental reversal of the right-aggregated \textit{l\textsubscript{B}} to the left-aggregated \textit{l\textsubscript{F}}. This incremental reversal is tantamount to adjusting pointers that causes shifts in sub-list boundaries. Thus, elements of \textit{aggs} may change from one sub-list to another and need to have its aggregation value recomputed.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/query-bloom.png}
      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

@@

\subsection*{False Positive rate}
% error bound related with parameters
The false-positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}. It is assumed that the hash functions produce an independent and random index value for each element.

Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula. The probability of one arbitrary bit being set to 1 is:
\begin{equation}
    \frac{1}{m}
\end{equation}
because there are \textit{m} bits. Likewise, the probability of one bit not being set to 1 is
\begin{equation}
    1-\frac{1}{m}
\end{equation}

Mullin states the probability that an arbitrary bit is not set to 1 after one element insertion and corresponding \textit{k} bit updates is:
\begin{equation}
    (1-\frac{1}{m})^\textit{k}
\end{equation}

For \textit{n} insertions, the probability a bit is not set to 1 is:
\begin{equation}
    (1-\frac{1}{m})^\textit{kn}
\end{equation}

Therefore, the probability that an arbitrary bit is set to 1 after \textit{n} insertions is:
\begin{equation}
    1-(1-\frac{1}{m})^\textit{kn}
\end{equation}

@@

Finally, this false positive rate can be approximated by 
\begin{equation}
    (1-e^\textit{-kn/m})^\textit{k}
\end{equation}

\subsection*{Optimal number of hash functions}
We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate. In order to do so, we must first allocate \textit{m} bits for the Bloom filter, choosing \textit{m} based on available memory. The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by:
\begin{equation}
    \frac{m}{n}ln(2)
\end{equation}

@@

Note that the time complexity does not at all depend on the number of elements processed by the Bloom filter and that \textit{k} will be a rather small constant. Hence, the Bloom filter is said to have constant time complexity.

@@

% data structures
\subsection*{Data structure}
A CMS is represented by a two-dimensional array (or matrix) with width \textit{w} and depth \textit{d}. A CMS will make use of \textit{d} hash functions, each associated with one row of the matrix. Each hash function will map elements to a position between 1 and \textit{w}, the column of the two-dimensional array. Initially, all of the matrix positions are 0. Figure \ref{fig:initial-cms} represents the two-dimensional array used by the CMS and corresponding dimensions in its initial state.

@@

% algorithm for insertion
\subsection*{Element Insertion}
When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, we hash the element using that row's hash function to obtain the result $j$. Lastly, for all obtained pairs of \textit{i} and \textit{j}, increment matrix cells \textit{(i, j)} value by 1, as seen in Figure \ref{fig:cms}. If elements have different weights these can be added instead of incrementing one unit.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/insertion-cms.png}
      \caption[Count-Min Sketch insertion]{CMS mapping element \textit{el} using \textit{d}=4 hash functions to determine the column position for each row and increment it.}
      \label{fig:cms}
    \end{center}
\end{figure}

% algorithm for query
\subsection*{Frequency Query}
Retrieving the minimum frequency of an element \textit{el} can be done by taking the minimum value of all row counts for \textit{el}. Mathematically, the frequency estimate is the minimum value of the set of all row values, for each row \textit{i} of the total \textit{d} rows, given by:
\[ min \{matrix[i][\textit{h\textsubscript{i}}(el)]\} \]

@@

\subsection*{Element Insertion}
In HyperLogLog, new elements are hashed into a binary string. The first \textit{n} bits of the hashed representation will map to one of the $m=2^n$ buckets. Effectively speaking, the first \textit{n} bits work as a \textit{bucket ID}. The bit-pattern observable \textit{P} --- the position of the leftmost 1-bit --- is counted from the binary string discarding the first \textit{n} bits. The observable \textit{P} is stored in the pre-determined bucket if it is greater than the value the bucket contains.

Figure \ref{fig:hll-insert} demonstrates such procedure for a new element \textit{el}. The hash representation of \textit{el}, using $n=2$ bits, maps to bucket \textit{10}. Discarding the first $n=2$ bits and counting from 1, the position of the leftmost 1-bit is $P=4$. The previous value of bucket \textit{10} was 0 and because $P=4$ is greater than 0 the bucket value is updated.
